---
title: "Chicago 311 Service Requests, in `{gt}` Tables"
author: "Jim Kloet (jim.kloet@gmail.com)"
date: "2020-10-31"
output:
  html_document:
    theme: flatly
    toc: true
    code_folding: "show"

---

<style type="text/css">

  body{
  font-size: 12pt;
  }
  
  <!-- uncomment below for full width tables (look too wide for my tastes) -->
  <!-- div.main-container { -->
  <!-- width: 100%; -->
  <!-- max-width: unset; -->
  <!-- } -->
  
</style>

```{r setup, include=FALSE}
# chunk options
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

### Overview

Tables are often treated as lesser forms of data visualization, but that is unfair: well-designed tables can support data-driven narratives, and look dapper in the process. Smarter people than me have laid out some principles for creating effective tables - see [Show Me the Numbers](https://www.amazon.com/Show-Me-Numbers-Designing-Enlighten/dp/0970601999) by Stephen Few, and [this 2020 paper](https://www.cambridge.org/core/journals/journal-of-benefit-cost-analysis/article/ten-guidelines-for-better-tables/74C6FD9FEB12038A52A95B9FBCA05A12) from Jonathan A. Schwabish (summarized on [this thread](https://twitter.com/jschwabish/status/1290323581881266177)), which include: 

* Adding a nice title and subtitle, to ensure that audiences know what a table is about even if it's shared separately from it's context
* Adding nice footnotes and source information, so the audience knows what the table means and where the data came from, without being overly distracted by expository text at the top of the table
* Easy to align columns (i.e. for right-aligning all numeric columns, ensuring consistent rounding of significant digits within columns), to facilitate visual scanning up and down the columns
* Easy to add row-striping, to facilitate visual tracking of values within row
* Minimizing non-data ink by using subtle divisions between cells and columns (if any!) and other non-obtrusive stylistic additions to the table

The `{gt}` package makes it easy to implement all of this in `R`, using the familiar syntax of the `tidyverse`. For a great set of tutorials, I recommend the 

I like working with Chicago 311 service request data, so this document makes some tables using those data, which all available from the [Chicago Data Portal](https://data.cityofchicago.org/Service-Requests/311-Service-Requests/v6vf-nfxy). The data in these tables includes service requests created on or between 31 Oct 2019 and 31 Oct 2020, and excludes the two most common requests, 311 information requests and aircraft noise complaints, because they are not associated with specific geographic locations in the available 311 request data: all information requests are associated with the main 311 office location, and all aircraft noise complaints are associated with airport locations.

```{r libraries and data}

library(tidyverse)
library(xml2)
library(rvest)
library(gt)

# read 311 data

raw_311 <- try(readRDS('./data/lastyear_311_requests.rds'), silent = TRUE)

if (class(raw_311) == 'try-error') {
  message('no local version of 311 data, grabbing from github...')
  raw_311 <- readRDS(
  url("https://github.com/jimtheflash/rstudio.tablecontest.2020/raw/main/data/lastyear_311_requests.rds","rb")
  )
}

# get CA lookup for merging
ca_lu <- read_html('https://en.wikipedia.org/wiki/Community_areas_in_Chicago') %>%
  html_table() %>%
  `[[`(1) %>%
  transmute(community_area = as.numeric(`Number[8]`),
            ca_name = `Name[8]`,
            ca_sq_mi = as.numeric(`Area (sq mi.)[10]`))

# read latest CA population data
raw_ca <- read_csv("https://github.com/jimtheflash/rstudio.tablecontest.2020/raw/main/data/ReferenceCCAProfiles20142018.csv") %>%
  transmute(ca_name = GEOG,
            ca_est_pop = floor(as.numeric(TOT_POP)))

# tidy and merge for tables
tidy_311 <- raw_311 %>%
  # get rid of the top 2 types of service requests
  filter(sr_type != '311 INFORMATION ONLY CALL',
         sr_type != 'Aircraft Noise Complaint') %>%
  mutate(grouping_sr_number = if_else(parent_sr_number == "", sr_number, parent_sr_number)) %>%
  select(grouping_sr_number,
         sr_number,
         sr_type,
         created_date,
         community_area) %>%
  group_by(grouping_sr_number) %>%
  mutate(distinct_requests_by_grouping_sr_number = n_distinct(sr_number)) %>%
  ungroup() %>%
  inner_join(ca_lu, by = 'community_area') %>%
  inner_join(raw_ca, by = "ca_name")
```

### Most Frequent Service Requests

This table summarizes the most frequent service requests by type, so we can identify the most frequent types of 311 service requests.

```{r most frequent requests}

tidy_311 %>%
  group_by(`Service Request Type` = sr_type) %>%
  summarise(`Total Requests` = n_distinct(sr_number),
            `Parent Issues` = n_distinct(grouping_sr_number)) %>%
  mutate(`Requests Per Parent Issue` = `Total Requests` / `Parent Issues`) %>%
  filter(`Total Requests` >= 10000) %>%
  arrange(desc(`Total Requests`)) %>%
  gt() %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = list(cells_title(),
                     cells_column_labels(columns = gt::everything())
  )) %>%
  tab_header(title = "Most Frequent 311 Service Requests in Chicago",
             subtitle = "Includes request types with at least 10,000 total requests from Oct 31 2019 through Oct 31 2020") %>%
  opt_align_table_header("left") %>%
  tab_footnote(footnote = "311 information requests and complaints about aircraft noise were excluded",
               locations = cells_title(groups = "subtitle")) %>%
  tab_footnote(footnote = "Total Requests is the number of requests filed, which can include requests about the same parent issue (e.g. multiple requests about a particular street light malfunctioning)",
               locations = cells_column_labels(vars(`Total Requests`))) %>%
  tab_footnote(footnote = "Parent Issues is the number of distinct issues of a given type (e.g. if there were 20 requests about 2 street lights malfunctioning, Parent Issues is 2)",
               locations = cells_column_labels(vars(`Parent Issues`))) %>%
  tab_source_note("311 service request data from https://data.cityofchicago.org/Service-Requests/311-Service-Requests/v6vf-nfxy") %>%
  fmt_number(columns = vars(`Total Requests`, `Parent Issues`),
             decimals = 0) %>%
  fmt_number(columns = vars(`Requests Per Parent Issue`),
             decimals = 3) %>%
  cols_align(align = 'right', columns = vars(`Total Requests`, `Parent Issues`, `Requests Per Parent Issue`)) %>%
  opt_row_striping()

```

$~$

Graffiti removal tops the list of Total Requests, as well as the list of Parent Issues (and by a wide margin!). The only request type on the list with more than two requests per parent issue was street light outages (which is why I used that example in the footnotes). I was also surprised to see that there was only one request per parent issue for abandoned vehicle complaints, which makes me think that that request means something very specific, but outside the scope of this writeup.

### Service Requests And Community Areas

This table summarizes 311 requests by community areas, so we can identify which community areas are submitting the most service requests. I thought total requests per person would be interesting as a way to gauge how frequently residents of an area were engaging with the 311 service, and I thought that parent requests per square mile might be an interesting way of estimating how many issues need addressing across community areas. This table is sorted by the Total Requests column.

```{r sr by ca}

tidy_311 %>%
  group_by(`Community Area` = ca_name) %>%
  summarise(`Est. Population` = max(ca_est_pop),
            `Area (Square Miles)` = max(ca_sq_mi),
            `Total Requests` = n_distinct(sr_number),
            `Parent Issues` = n_distinct(grouping_sr_number)) %>%
  mutate(`Requests Per Person` = `Total Requests` / `Est. Population`,
         `Parent Issues Per Sq. Mi.` = `Parent Issues` / `Area (Square Miles)`) %>%
  filter(`Total Requests` >= 15000) %>%
  arrange(desc(`Total Requests`)) %>%
  gt() %>%
  tab_style(
    style = cell_text(size = 'small'),
    locations = list(cells_column_labels(columns = gt::everything()),
                     cells_body())
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = list(cells_title(),
                     cells_column_labels(columns = gt::everything())
  )) %>%
  tab_header(title = "311 Service Requests By Chicago Community Area",
             subtitle = "Includes community areas with at least 15,000 total requests from Oct 31 2019 through Oct 31 2020") %>%
  opt_align_table_header("left") %>%
  tab_footnote(footnote = "311 information requests and complaints about aircraft noise were excluded",
               locations = cells_title(groups = "subtitle")) %>%
  tab_footnote(footnote = "Total Requests is the number of requests filed, which can include requests about the same parent issue (e.g. multiple requests about a particular street light malfunctioning)",
               locations = cells_column_labels(vars(`Total Requests`))) %>%
  tab_footnote(footnote = "Parent Issues is the number of distinct issues of a given type (e.g. if there were 20 requests about 2 street lights malfunctioning, Parent Issues is 2)",
               locations = cells_column_labels(vars(`Parent Issues`))) %>%
  tab_source_note("Estimated community area population taken from American Community Survey 2014-2018 via https://datahub.cmap.illinois.gov/dataset/community-data-snapshots-raw-data/resource/8c4e096e-c90c-4bef-9cf1-9028d094296e") %>%
  tab_source_note("Community area square mileage taken from https://en.wikipedia.org/wiki/Community_areas_in_Chicago") %>%
  tab_source_note("311 service request data from https://data.cityofchicago.org/Service-Requests/311-Service-Requests/v6vf-nfxy") %>%
  tab_options(footnotes.font.size = "small",
              source_notes.font.size = "small") %>%
  fmt_number(columns = vars(`Est. Population`, `Total Requests`, `Parent Issues`, `Parent Issues Per Sq. Mi.`),
             decimals = 0) %>%
  fmt_number(columns = vars(`Area (Square Miles)`),
             decimals = 2) %>%
  fmt_number(columns = vars(`Requests Per Person`),
             decimals = 3) %>%
  cols_align(align = "right", 
             columns = vars(`Area (Square Miles)`, `Est. Population`, `Total Requests`, `Parent Issues`, `Requests Per Person`,`Parent Issues Per Sq. Mi.`)) %>%
  opt_row_striping()

```

$~$

This table, despite being fairly long, is still pretty digestible. While I think the sorting (Total Requests) is sensible, it is also feasible that readers would want to sort by other criteria. In these cases, a table made with the `{reactable}` or `{DT}` libraries may be a better option, but comes with limitations: neither of those libraries enable nice titles, or the footnotes and source information found here. If the information found in those sections can be added elsewhere, then those options might be better for these use cases.

(And obviously an interactive map would be a reasonable way to visualize these data!)

### Top Service Requests by Community Area

This table summarizes the top five service requests types by community area, sorted by community area population. This makes it possible to identify community areas with unique mixes of service request types.

```{r interaction table}

tidy_311 %>%
  group_by(ca_name) %>%
  filter(n_distinct(sr_number) >= 15000) %>%
  ungroup() %>%
  group_by(`Community Area` = ca_name, `Service Request Type` = sr_type) %>%
  summarise(pop = max(ca_est_pop),
            `Total Requests` = n_distinct(sr_number),
            `Parent Issues` = n_distinct(grouping_sr_number)) %>%
  arrange(desc(pop), desc(`Total Requests`)) %>%
  mutate(rank = row_number()) %>%
  filter(rank <= 5) %>%
  select(-pop, -rank) %>%
  gt() %>%
  tab_style(
    style = cell_text(size = "medium"),
    locations = cells_title()
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = list(cells_title(),
                     cells_column_labels(columns = gt::everything()),
                     cells_row_groups()
  )) %>%
  tab_style(
    style = cell_fill(alpha = 0.5),
    locations = cells_row_groups()
  ) %>%
  tab_header(title = "Top 311 Service Request Types by Chicago Community Area",
             subtitle = "Includes community areas with at least 15,000 total requests from Oct 31 2019 through Oct 31 2020") %>%
  opt_align_table_header("left") %>%
  tab_footnote(footnote = "311 information requests and complaints about aircraft noise were excluded",
               locations = cells_title(groups = "subtitle")) %>%
  tab_footnote(footnote = "Total Requests is the number of requests filed, which can include requests about the same parent issue (e.g. multiple requests about a particular street light malfunctioning)",
               locations = cells_column_labels(vars(`Total Requests`))) %>%
  tab_footnote(footnote = "Parent Issues is the number of distinct issues of a given type (e.g. if there were 20 requests about 2 street lights malfunctioning, Parent Issues is 2)",
               locations = cells_column_labels(vars(`Parent Issues`))) %>%
  tab_source_note("Estimated community area population taken from American Community Survey 2014-2018 via https://datahub.cmap.illinois.gov/dataset/community-data-snapshots-raw-data/resource/8c4e096e-c90c-4bef-9cf1-9028d094296e") %>%
  tab_source_note("Community area square mileage taken from https://en.wikipedia.org/wiki/Community_areas_in_Chicago") %>%
  tab_source_note("311 service request data from https://data.cityofchicago.org/Service-Requests/311-Service-Requests/v6vf-nfxy") %>%
  tab_options(footnotes.font.size = "small",
              source_notes.font.size = "small") %>%
  fmt_number(columns = vars(`Total Requests`, `Parent Issues`),
             decimals = 0) %>%
  cols_align(align = "right", columns = vars(`Total Requests`, `Parent Issues`))

```
$~$

This is a lot cleaner than I expected it to be! This table is very long, even after filtering to a subset of community areas; it would be nice to be able to page between community area lists (with may be feasible with `{reactable}` or `{DT}`), but for a website or blog this formatting is still digestible, I think.

### Summary

Building well-designed tables in `R` is easy with the `tidyverse` and `{gt}` library. 
